---
title: "Supplementary Material"
subtitle: "Summary of Results of Plotkin et al 2022"
author: "Aaron R. Caldwell"
format: 
  pdf:
    toc: true
    link-citations: true
execute:
  warning: false
  messages: false
  echo: false
editor: source
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

\newpage

# Scripts and Accessibility

This reproducible analysis document runs source scripts. These can be found on the independent, version-controlled, data repository for this project.

```{r}

source(
  here::here("scripts",
  "plotkin_analysis_example.R")
)
```

# Introduction

This document is a short summary of the re-analysis of the 1-RM back squat outcome that was reported in the manuscript of @plotkin2022. For this re-analysis, we replicate the model presented in the manuscript wherein the change score was the outcome, the independent variable was the training group (LOAD vs REPS), and two covariates (sex and pre-intervention 1-RM) were included in the model. The results slightly differ because we do not use the bootstrapping method mentioned in the original manuscript.

We present the average treatment effect using the estimated marginal means using the `emmeans` R package [@emmeans]. Following this, we present various tests of heteroskedasticity and the bounds for the heterogeneity of treatment effects [@gadbury2001].

\newpage

# Summary Statistics

This match the publication, barring one SD which appears to be a typo in the original publication[^1]

[^1]: The delta in the LOAD group is incorrectly presented as 21.2 rather than 12.2 in Table 1 of @plotkin2022

```{r}
sum_table %>%
  as_flex_table() %>%
  flextable::theme_apa()
```

## Plot of Change Scores

```{r}
plot_delta
```

\newpage

# Model Summary

Note: pre-scores not centered or scaled.

```{r}
model %>%
  as_flextable()
```

\newpage

# Average Treatment Effect

The average treatment effect can be provided by a contrast using the estimated marginal means (`emmeans` package). The differences differ slightly from the publication because we utilize a parametric model while the original authors calculated confidence intervals using bootstrap methods.

```{r}
ATE %>%
  tibble() %>%
  flextable()
```

\newpage

# Tests for Heteroskedasticity

The results of the different tests are mostly in agreement. However, note that the $s_{IR}$ (last test) is negative which, under the logic of $s_{IR}$ proponents, would end any further investigation of heterogeneity of treatment effects. Additionally, none of the tests are "significant", but such tests would be quite underpowered given the study's sample size.

## Breusch-Pagan Test (Model Residuals)

```{r}
bp_test
```

## Variance Ratio Test

```{r}
var_ratio_test
```

\newpage

## Log Variability (SD) Test

```{r}
lnvar_test
```

## Standard Deviation of Individual Response

```{r}
sd_ir
```

\newpage

# Bounds on HTE

We can then use the work of @gadbury2001 to get the bounds on standard deviation of the treatment effect. A figure and table of the results are below. The average treatment from the model was utilized to calculate the $P_{harmed}$. Also, we opted to use the method of creating bounds that does *not* utilize covariate information, but the code to do such an analysis is included on our independent repository.

## Method to calculate $s_D$ and its confidence intervals:

Let:

- $n_x$ = size of treatment group
- $n_y$ = size of control group
- $\hat{\sigma}_x$ = sample standard deviation of treatment group
- $\hat{\sigma}_y$ = sample standard deviation of control group
- $\rho_{xy}$ = correlation between treatment and control measurements

The standard deviation of individual responses ($\hat{\sigma}_D$) is estimated as follows:

$\hat{\sigma}_D^2 = \hat{\sigma}_x^2 + \hat{\sigma}_y^2 - 2\hat{\sigma}_x\hat{\sigma}_y\rho_{xy}$

The variance of this estimate is given by:
$Var(\hat{\sigma}_d^2) = 2\left[\frac{\hat{\sigma}_x^2}{n_x}(\hat{\sigma}_x - \rho_{xy}\hat{\sigma}_y)^2 + \frac{\hat{\sigma}_y^2}{n_y}(\hat{\sigma}_y - \rho_{xy}\hat{\sigma}_x)^2\right]$

For a confidence level of $1-\alpha$, the confidence interval is constructed as:

$\hat{\sigma}_D^2 \pm z_{1-\alpha/2}\sqrt{Var(\hat{\sigma}_d^2)}$

where $z_{1-\alpha/2}$ is the critical value from the standard normal distribution.

The final estimate of the standard deviation of individual responses and its confidence interval are obtained by taking the square root:
$\hat{\sigma}_D = \sqrt{|\hat{\sigma}_D^2|} \cdot sign(\hat{\sigma}_D^2)$

For the confidence interval bounds:

$CI_{lower,upper} = \sqrt{|\hat{\sigma}d^2 \pm z{1-\alpha/2}\sqrt{Var(\hat{\sigma}_d^2)}|} \cdot sign(\hat{\sigma}d^2 \pm z{1-\alpha/2}\sqrt{Var(\hat{\sigma}_d^2)})$

\newpage

## Method to calculate $P_{harmed}$ and its confidence intervals:

Let's define the additional variables:

- $\hat{\mu}_D$ = estimated mean treatment effect
- $\hat{\sigma}_D$ = estimated standard deviation of individual responses
- $\phi(\cdot)$ = standard normal probability density function
- $\Phi(\cdot)$ = standard normal cumulative distribution function

The probability of harm (unfavorable effect) is estimated as:

$\hat{P}_{harmed} = \Phi\left(\frac{\hat{\mu}_D}{\hat{\sigma}_D}\right)$

The variance of this estimate requires several components.

First, the variance of the mean treatment effect:

$Var(\hat{\mu}_D) = \frac{\hat{\sigma}_x^2}{n_1} + \frac{\hat{\sigma}_y^2}{n_2}$

Then, using the delta method, the variance of $\hat{P}_{-}$ is:

$Var(\hat{P}_{harmed}) = \frac{\phi^2(\hat{\mu}_D/\hat{\sigma}_D)}{\hat{\sigma}_D^2}\left[Var(\hat{\mu}_D) + \frac{\hat{\mu}_D^2 \cdot Var(\hat{\sigma}_D^2)}{4\hat{\sigma}_D^4}\right]$

where $\phi(\hat{\mu}_D/\hat{\sigma}_D)$ is the standard normal density evaluated at $\hat{\mu}_D/\hat{\sigma}_D$

For a confidence level of $1-\alpha$, the confidence interval is constructed as:

$\hat{P}_{harmed} \pm z_{1-\alpha/2}\sqrt{Var(\hat{P}_{harmed})}$

where $z_{1-\alpha/2}$ is the critical value from the standard normal distribution.
This formulation provides an estimate of the probability that an individual will experience an unfavorable response to treatment, along with uncertainty bounds around that estimate.

*Note*: The confidence interval bounds should be constrained to [0,1] since we are estimating a probability. However, as we can see in the visualizations, with small sample sizes this approach can produce bounds outside the interval.

\newpage

## Visualization of $s_D$ and $P_{harmed}$

```{r}
#| fig-height: 7
#| fig-width: 6.5
p_final
```

\newpage

## Table of Bounds

```{r}

var_d_mle %>%
  flextable() %>%
  flextable::padding(padding.bottom = 0,padding.top = 0, part = "all") %>%
  line_spacing(part = "body", space = 1.1) %>%
  theme_apa()  %>% 
  fontsize(part = "all",
       size =8)
```

\newpage

# References
